Structure Learning for $VAR(p)$ Models

### 2019-03-10 paper/paper.tex

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{P(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node

@@

The equivalence $(1) \iff (3)$ is essentially a restatement of the
definition, but is in line with the seminal work of Geweke
\cite{geweke1982}, \cite{geweke1984}.  Statement (4) means that
$(y - \linE{y}{\mathcal{H}_{t - 1}^{-j}}) \perp (z -
\linE{z}{\mathcal{H}_{t - 1}^{-j}})\ \forall y \in \mathcal{H}_t^i,\ z \in
\mathcal{H}_{t - 1}^j$, which is an appealing geometric statement that
there is no direct feedback from $x_j$ to $x_i$ \cite{lindquist}.

@@

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  That
is, $x_k$ may be a common ancestor that when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality, where $x_i \pwgc x_j$ if
$x_i$ Granger-causes $x_j$ with respect only to $(x_i, x_j)$.

@@

@@

Finally, the \textit{propagation delay} from $j$ to $i$, denoted $d(j, i)$, will be defined as

\[
  d(j, i) = \text{min}\big\{\sum_{k = 0}^{K - 1} d(p_k, p_{k + 1}})\ |\ i = p_0 \rightarrow p_1 \rightarrow \cdots \rightarrow p_K = j \text{ is a path in } \gcg \big\}
\]

and $d(j, i) \defeq \text{min}\{B_{i, j}(\tau) \ne 0\}$ if $j \in \pa{i}$.

\begin{lemma}
  \label{lem:pwgc_anc}
  $x_j \pwgc x_i$ if and only if at least one of the following hold

  \begin{enumerate}
    \item{$j \in \anc{i}$}
    \item{$\anc{i}\cap\anc{j} \ne \emptyset$.} %and $d(k, j) < d(k, i)$}
  \end{enumerate}
\end{lemma}

@@

\begin{proof}
\end{proof}

@@

\begin{theorem}[Tree Recovery]
  \label{thm:tree_recovery}
  If the Granger-Causality graph $\gcg$ for $x(t)$ is a
  directed forest, then $\gcg$ can be inferred from pairwise
  causality tests alone.
\end{theorem}

>>>>

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{P(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\gcgpath}[4]{#1 \rightarrow #2_1 \rightarrow \cdots \rightarrow #2_{#3} \rightarrow #4}

@@

The equivalence $(1) \iff (3)$ is essentially a restatement of the
definition, but is in line with the seminal work of Geweke
\cite{geweke1982measurement}, \cite{geweke1984}.  Statement (4) means that
$(y - \linE{y}{\mathcal{H}_{t - 1}^{-j}}) \perp (z -
\linE{z}{\mathcal{H}_{t - 1}^{-j}})\ \forall y \in \mathcal{H}_t^i,\ z \in
\mathcal{H}_{t - 1}^j$, which is an appealing geometric statement that
there is no direct feedback from $x_j$ to $x_i$ \cite{lindquist}.

@@

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality, where $x_i \pwgc x_j$ if
$x_i$ Granger-causes $x_j$ with respect only to $(x_i, x_j)$.

@@

% What if we take this as the definition of PWGC?  then, in a simply causal graph
% the "normal" definition becomes a theorem.  Or, if lem:pwgc_anc is true then
% maybe that is good enough?

\begin{lemma}
  \label{lem:pwgc_anc}
  If $j \in \anc{i}$ then $j \pwgc i$.
\end{lemma}
\begin{proof}
  This is actually not true.  e.g.  $x_j(t) = v_i(t) + x_k(t - 1) + x_{k'}(t - 1)$ with $x_k(t) = v_k(t) + ax_j(t - 1)$ and $x_{k'}(t) = v_{k'}(t) - ax_j(t - 1)$.
\end{proof}

@@

% Finally, the \textit{propagation delay} from $j$ to $i$, denoted $d(j, i)$, will be defined as

% \[
%   d(j, i) = \text{min}\big\{\sum_{k = 0}^{K - 1} d(p_k, p_{k + 1}})\ |\ i = p_0 \rightarrow p_1 \rightarrow \cdots \rightarrow p_K = j \text{ is a path in } \gcg \big\}
% \]

% and $d(j, i) \defeq \text{min}\{B_{i, j}(\tau) \ne 0\}$ if $j \in \pa{i}$.

\begin{lemma}
  \label{lem:pwgc_anc}
  If $x_j \pwgc x_i$ then one or more of the following hold

  \begin{enumerate}
    \item{$j \in \anc{i}$}
    \item{$\anc{i}\cap\anc{j} \ne \emptyset$.}
  \end{enumerate}
\end{lemma}

@@

\begin{proof}
  Is this true if we take \ref{lem:pwgc_anc} as an axiom?
\end{proof}

@@

We are interested in understanding under what conditions a causality
graph can be correctly recovered by pairwise testing alone.  It is not
enough that the graph simply be sparse, as the following example shows.

\begin{example}
  If the causality graph $\gcg$ of the vector process $x(t)$ is
  strongly connected\footnote{there is a directed path from each node
    to every other node}, then the causality graph recovered via
  pairwise causality tests is the complete graph.

  This follows immediately from the fact (lemma \ref{lem:pwgc_anc})
  that $j \in \anc{i} \implies j \pwgc i$.
\end{example}

\begin{definition}[Simply Causal]
  If the Granger-Causality graph $\gcg$ for $x(t)$ is as a directed
  acyclic graph such that there is at most one non-overlapping
  directed path between any two nodes (excluding self-loops) then we
  will refer to $\gcg$ as being \textit{simply causal}.
\end{definition}

\begin{example}
  A directed tree (more generally directed forest) is a simply causal
  graph.
\end{example}

\begin{lemma}
  If $\gcg$ is simply causal then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \implies j \notin \anc{i}$.
\end{lemma}
\begin{proof}
  Suppose that $\gcg$ is simply causal and that we have both
  $i \pwgc j$ and $j \in \anc{i}$.  We will show that $i \in \anc{j}$,
  which produces a contradiction (since $\gcg$ being a DAG is part of
  the requirements for being simply causal).

  Since $j \in \anc{i}$ there is a path $\gcgpath{j}{a}{p}{i}$.
  Moreover, since $i \pwgc j$ by lemma \ref{lem:pwgc_anc} there must
  be a common ancestor $k \in \anc{i} \cap \anc{j}$ (otherwise, we would
  have $i \in \anc{j}$, contradicting the acyclic property).  The node
  $k$ has paths to both $j,\ \gcgpath{k}{b}{q}{j}$ and $i,\ \gcgpath{k}{c}{r}{i}$.

\begin{theorem}[Pairwise Recovery]
  \label{thm:tree_recovery}
  If the Granger-Causality graph $\gcg$ for $x(t)$ is simply causal
  then then $\gcg$ can be inferred from pairwise causality tests
  alone.
\end{theorem}
\begin{proof}

\begin{example}
  The condition of simple causality is merely a sufficient condition.  For example, the complete directed graph with 2 nodes (e.g. $B(1) = \left[ \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right]$) contains a loop but is pairwise recoverable.
\end{example}
